First paper - Draft

### 2017-07-05 paper/paper.tex
Fixed table and graphs. Everything ready.

It is incomplete to tusy educational inequalities without considering the previlance of intergenerational transmission of human and social capital. Social mobility has deepend our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity. The field of economics of education has also helped to understand that education is by far the investment that yields the best returns. This is not only for the child and his/her family, but to society as a whole, as it boosts economic activity, it helps the labour market to improve job conditions and maintain a rapid economic growth \citep{hanushek2007}. We've become so intertwined with the subject that we've virtually spent most of our time studying the mechanisms through which this inequality comes about. Naturally, we want to do that in order to reverse it and help every child reach its fullest potential. Despite our good intentions, we've concentrated little on the magnitud of the gaps, specially in terms of cognitive abilities. We want to know if the gap is big or small relative to other countries. Furthermore we want to know if this gap has reponded to our policy efforts to reduce it.

@@

% Show the properties of the ESCS dummy variable
% % in high educated
% % of home possessions
% % of ISEI

% Figure out how to ouput data frames as latex tables.

% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Wed Jul  5 09:41:15 2017
\scalebox{0.75}{
\begin{tabular}{llllllll}
   \hline \
                          & & & Low SES & & & High SES \\
                          \cmidrule(l){3-5}
                          \cmidrule(l){6-8} Year & Countries & \% Low Edu & \% High Edu & Sample size & \% Low Edu & \% High Edu & Sample size \\ \hline2000 & Australia & 5\% & 1\% & 294 & 0\% & 97\% & 270 \\ 
  2000 & Austria & 4\% & 0\% & 294 & 0\% & 99\% & 235 \\ 
  2000 & Canada & 3\% & 1\% & 1970 & 0\% & 99\% & 1357 \\ 
  2000 & Denmark & 4\% & 0\% & 230 & 0\% & 97\% & 228 \\ 
  2000 & Finland & 2\% & 0\% & 261 & 0\% & 97\% & 267 \\ 
  2000 & France & 14\% & 0\% & 253 & 0\% & 97\% & 258 \\ 
  2000 & Germany & 2\% & 0\% & 250 & 0\% & 93\% & 303 \\ 
  2000 & Hungary & 1\% & 0\% & 291 & 0\% & 99\% & 254 \\ 
  2000 & Italy & 3\% & 0\% & 278 & 0\% & 99\% & 278 \\ 
  2000 & Netherlands & 9\% & 0\% & 129 & 0\% & 99\% & 158 \\ 
  2000 & Poland & 0\% & 0\% &  & 0\% & 92\% & 166 \\ 
  2000 & Spain & 15\% & 0\% & 318 & 0\% & 98\% & 340 \\ 
  2000 & Sweden & 4\% & 1\% & 241 & 0\% & 98\% & 246 \\ 
  2000 & United Kingdom & 2\% & 0\% & 531 & 0\% & 95\% & 415 \\ 
  2000 & United States & 5\% & 0\% & 222 & 0\% & 97\% & 164 \\ 
  2015 & Australia & 2\% & 2\% & 1694 & 0\% & 94\% & 1235 \\ 
  2015 & Austria & 4\% & 0\% & 676 & 0\% & 96\% & 705 \\ 
  2015 & Canada & 1\% & 3\% & 2215 & 0\% & 97\% & 1863 \\ 
  2015 & Denmark & 4\% & 0\% & 1013 & 0\% & 99\% & 595 \\ 
  2015 & Finland & 2\% & 1\% & 575 & 0\% & 99\% & 584 \\ 
  2015 & France & 3\% & 0\% & 570 & 0\% & 94\% & 615 \\ 
  2015 & Germany & 5\% & 0\% & 545 & 0\% & 96\% & 582 \\ 
  2015 & Hungary & 1\% & 0\% & 466 & 0\% & 99\% & 589 \\ 
  2015 & Italy & 2\% & 0\% & 947 & 0\% & 96\% & 1043 \\ 
  2015 & Netherlands & 7\% & 0\% & 521 & 0\% & 83\% & 525 \\ 
  2015 & Poland & 0\% & 0\% &  & 0\% & 97\% & 448 \\ 
  2015 & Spain & 10\% & 0\% & 608 & 0\% & 99\% & 703 \\ 
  2015 & Sweden & 7\% & 2\% & 527 & 0\% & 97\% & 542 \\ 
  2015 & United Kingdom & 2\% & 1\% & 1387 & 0\% & 93\% & 1388 \\ 
  2015 & United States & 10\% & 0\% & 585 & 0\% & 97\% & 544 \\ 
   \hline
\end{tabular}
}

\subsection{Results}

Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).

@@

We start by look at the achievement for some countries. We plot only mathematics for each country and also a quadratic trend spline for both test scores pooled. Some countries have increased their achievement very strongly. For example, France, Austria and surprisingly Sweden. In mathematics, France increase the gap by roughly 0.9. This pattern happens similarly for literacy with an increase of 0.6. For such a short period of time, the magnitud of the increase is reasonably big.

Given that no one has estimated the evolution of the gap we can't cross-check how other empirical estimations put France at. However, we can make some educated inferences.  For example, \citet{micklewright} also finds that France is a low dispersion country, specifically in the year 2000.

However, we don't know what happens over time. But \citet{bernardi2016} did study the relationship of social origin effects and found that they increased very much in France in the last decades.

Other countries have reasonable increases such as Austria and Finland. Aside from these countries, some other countries have no change at all. Specially, Canada, Netherlands and Spain. Canada excels here not only because the gap has been stable over time, but because it has the smallest gap of all countries presented here. It's nearly 0.5 standard deviations in 2000 and it increased only by 0.2 in 2015. We can't really say Italy has had a significant since it has a sort of wiggly pattern that looks like it cancels itself out.

Furthermore, we do have some countries which show a decrease in the SES achievement gap. United Kingdom, Israel, Poland, United States and Germany show some steep downward slopes. Most notably are the U.S and German cases. These two countries show a very high level of dispersion in the year 2000 with SES gaps of over 2 standard deviations. However, both countries reduced the gaps by -0.6 and -0.9 respectively. Their distinctively big gaps in 2000 also show up in the work of \citet{micklewright}.

@@

The plot below takes a more direct approach and looks at the percentage change from the first and last timepoint available. Each data point has been computed together with it's 50\% and 95\% uncertainty interval \footnote{Each of these uncertainty intervals were computed using a 500-replicate bootstrap}.

@@

Generally speaking, we see that most countries increased their achievement gap. France had an average increase of about 80\% since 2000, whereas Germany had a similar figure but decreasing. Many of the countries we saw before that didn't have a very steep slope, such as Hungary or Australia, in fact had increases of about 40\% of their gap. In contrast, we see that the U.S and Poland had also quite significant decreases of about 40\%.

Something reassuring is that mathematics and reading follow basically the same trend across all countries \footnote{Reading had an even bigger decrease for similar countries, and a much smaller increase. The previos two graphs for reading are available upon request}. This means that the result is not an artifact of chance. However, it's important to disentangle where is this gap originating from. Is this because the top are improving why the bottom decreases? Or is it that the bottom is catching up? Up next we plot the same graph but show the divergent patterns between high/low SES origins.

@@

The top line belongs to the high SES group whereas the bottom line belongs to the low SES group. The middle line is a summary line showing whether one of the groups is growing/shrinking faster than the other. We find plenty of variation in these patterns. For example, in the United States the top seems to be equalizing much stronger than than the bottom is catching up. The U.K seems to be following the same path as the United States as well. On the other hand, in Poland the gap seems to be closing because the bottom SES group is catching up much faster. The Netherlands shows a similar pattern as the U.S, U.K patterns but upon closer inspection the explanation is very different. We see that the slope is negative (like in the previous two countries) but that's because the low SES group is decreasing at a greater rate than the high SES is decreasing.

Finally, we pay attention to the flat summary lines. These lines should be interpreted with caution because it doesn't mean that the gap is not increasing. Denmark, Belgium, Netherlands, Norway and Spain show a flat line because the gap is growing very little, if at all. On the other hand, Sweden, France, Finland, New Zealand, Austria, among other countries, show a flat line because both groups are distancing themselves at a very similar rate. there's a group of distinctive groups where both low/high SES rates are very great resulting in the summary line being flat.

Ths source of this gap can be deconstructed into two broader gaps. We want to know if these gaps have increased because the 90th SES group has distanced itself from the 50th SES group or because the middle has been also distancing itself from 10th SES group. Up next we plot all the combinations of the 90th/50th/10th gaps and pay attention to the magnitude of each gap. For example, we're interested in searching which gap is bigger between countries, the 90th/50th or the 50th/10th? Moreover, are some of these gaps contributing more to the 90th/10th gap we saw earlier? Because we visualize three plots for every country, we only plot a selected number of countries that represent the overall patterns from the previous graphs \footnote{United States and Germany show a decrease, France shows a steep increase and Denmark shows no significant change.}.

% This plot might not be necessary any more
% <<>>=
% 
% lower <- seq(13, 1, -1)
% upper <- seq(100, 220, 10)
% 
% df <- tibble(country = "fake_country",
%              years = rep(seq(2000, 2025, 2), 2),
%              ses = rep(c("lower", "upper"), each = 13),
%              value = c(lower, upper))
% 
% # You left off here. Plot the gap for 1 and 0 for all countries in the graph before.
% df %>%
%   ggplot(aes(as.factor(years), value, group = ses, colour = ses)) +
%   geom_point(size = 0.5, alpha = 0.4) +
%   geom_line(alpha = 0.4) +
%   geom_line(stat = "smooth", method = "lm", aes(group = 1),
%             formula = y ~ splines::ns(x, 3), linetype = "longdash",
%             colour = "black") +
%   # geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1, alpha = 0.4) +
%   # geom_hline(yintercept = 0, linetype = "longdash") +
%   # coord_cartesian(ylim = c(0, 3)) +
%   facet_wrap(~ country)
% @

@@

We start off with Denmark. Generally speaking, we find that the 50th/10th gap is contributing more than the 90th/50th gap as the gap is bigger for both tests. Furthermore, we find, just as in the 90th/10th SES gap, that there isn't a significant decrease, with the exception of reading in the 50th/10th SES gap which seemed to decrease from nearly 1.5 to 1 standard deviation. France shows a similar pattern as Denmark in that the 50th/10th SES gap is the main contributer and that the three gaps have the similar trend.

Germany and U.S, on the other hand, show a very different landscape. For example, Germany, shows that the 90th/10th gap is primarily driven by the 50th/10th SES gap and that the slopes of the 90th/50th gap is much weaker than the other two gaps. This means that Top/Middle SES groups have shrunk much less than the other groups. The big catching up is actually coming from the Middle/Bottom SES groups. However, we don't know if this is because the bottom is catching up or that the middle is coming down. However, based on the gap plots from above, we can infer that it is because the bottom is catching up rapidly because against the 90th SES group, it is increasing steeply every year. The U.S shows a different pattern from all other countries. Both the 90th/50th/10th gaps seem to contribute rather similarly to the overall 90th/10th gap (although 50th/10th is marginally higher) but more interestingly, the 90th/50th gap has stagnated in the last 15 years. The top/middle groups have a relatively big gap and it hasn't changed at all. On the other hand, the steep decrease we saw for the U.S is coming exclusively from the 50th/10th gap shrinking. The reason, nonetheless, might be different from the German case, where it's probable that the bottom is catching up. Here, as in the 90th/10th gap, the middle might be 'dumbing' down closer to the low SES group.

In a more convenient way, we can inspect whether there is a lot of variance in the rate at which top/bottom groups are changing.

@@

We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (red panel) but there's also a fair share where low SES is catching up faster, such as United States and Germany. We see a negative relationship where the faster the low SES catch up the slower the High SES decrease. Quite an interesting pattern because we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel). On the other hand, we do see more ocuntries where the both groups are increasing (top right panel). We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.

But, similar as before, we don't know which group is causing the decrease/increase of the gap in the 90th/50th and 50th/10th gap. For that we plot the evolution of these gaps below.

@@

Given that we have the data for the average performance for each country and the achievement gap, we can also investigate whether there's an efficiency/equality tradeoff and if it stands over time. One particular topic that has been researched in the past is whether improving the overall cognitive development of a country compromises the distribution of the countries overall performance. For example, in order to improve the average score in, let's say, mathematics, tracking can improve the average score by making children in the higher trackers perform better, raising the overall performance. This, however, can come at a price. If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied with an increase in the overall performance.

@@

{\centering \includegraphics[width=8.5in,height=5in]{figure/avgperformance_inequality-1}

@@

For the past 15 years we see a very strong correlation suggesting that countries that perform very well in PISA have actually very low achievement gaps (or the other way around) \footnote{We graph all countries-years pooled as well and the relationship is identical}. The correlations are between -0.7343267 and -0.8781625. This confirms the theoretical and empirical frameworks suggested by \citet{werfhorst_mijs}. However, what they didn't really go in depth was whether this corelation is strongest for some countries over time. With the data available we can look at the evolution separately for many countries.

@@

In fact, we see a strong correspondence. In countries like France we see quite a functional form between both inequality and performance. As average performance plummets, the achievement gap is increasing. This pattern is also seen in in the Austria, Australia and Germany. The other obvious pattern is the opposite trend. We see countries like Poland, Finland, Canada and much less pronounced Italy that this pattern is present. The surprising case here is the United States. We see that as the average performance has been decreasing, the average achievement gap has been decreasing as well. In fact, among the few countries we should find something that resembles this pattern should be the U.S because it doesn't have a strict institutionalized tracking system. However, the explanation behind this pattern is very speculative. Finally, the last notable pattern is that in some countries the average performance is so high that it actually overperforms the achievement gap. Only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality.

>>>>

It is incomplete to study educational inequalities without considering the prevalence of intergenerational transmission of human and social capital. Social mobility has deepend our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity. The field of economics of education has also helped to understand that education is by far the investment that yields the best returns. This is not only for the child and his/her family, but to society as a whole, as it boosts economic activity, it helps the labour market to improve job conditions and maintain a rapid economic growth \citep{hanushek2007}. Since \citet{coleman1966} we've spent most of our time studying the mechanisms through which this inequality comes about. Naturally, we want to do that in order to reverse it and help every child reach its fullest potential. Despite our good intentions, we've concentrated little on the magnitud of the gaps, specially in terms of cognitive abilities. We still don't know, in comparative terms, which countries have big or small cognitive gaps resulting from SES origins. And even more importantly, we haven't assesed whether policy efforts to reduce inequality have actually had an impact in reducing the achievement gap over time.

@@

% <<table_escs, include = TRUE>>=
% summary_table <-
%   left_join(summary_data, sample_tables_topbottom) %>%
%   filter(name %in% c("2000", "2015")) %>%
%   arrange(country, name)
% 
% table_coming <-
%   summary_table %>%
%   select(-perc) %>%
%   gather(cats, vals, -(name:escs_dummy)) %>%
%   unite(all_vals, escs_dummy, cats, sep = "_") %>%
%   spread(all_vals, vals) %>%
%   map_at(c(3, 4, 6, 7), ~ {
%     .x[is.na(.x)] <- 0
%     paste0(round(.x, 0), "%")
%   }) %>%
%   map_if(is_double, as.character) %>%
%   data.frame() %>%
%   xtable::xtable()
% 
% addtorow <- list()
% addtorow$pos <- list(0)
% addtorow$command <- paste("\\hline \\
%                           & & & Low SES & & & High SES \\\\
%                           \\cmidrule(l){3-5}
%                           \\cmidrule(l){6-8}",
%                           paste0(
%                             c("Year",
%                               "Countries",
%                               "\\% Low Edu",
%                               "\\% High Edu",
%                               "Sample size",
%                               "\\% Low Edu",
%                               "\\% High Edu",
%                               "Sample size"),
%                             collapse = " & "),
%                           "\\\\ \\hline")
% 
% print(table_coming,
%       scalebox = '0.75',
%       floating = FALSE,
%       add.to.row = addtorow,
%       include.colnames = FALSE,
%       include.rownames = FALSE,
%       hline.after = nrow(table_coming),
%       caption.placement = "top",
%       caption = "SES sample size and ISCED composition")
% @

% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Wed Jul  5 17:51:02 2017
\scalebox{0.75}{
\begin{tabular}{lllrrlrr}
   \hline \
                          & & & Low SES & & & High SES \\
                          \cmidrule(l){3-5}
                          \cmidrule(l){6-8} Year & Countries & N & Avg score & S.E & N & Avg score & S.E \\ \hline2000 & Australia & 294 & 0.22 & 0.07 & 270 & 1.22 & 0.09 \\ 
  2000 & Austria & 294 & -0.04 & 0.11 & 235 & 1.15 & 0.15 \\ 
  2000 & Canada & 1970 & 0.30 & 0.05 & 1357 & 0.90 & 0.07 \\ 
  2000 & Denmark & 230 & -0.01 & 0.13 & 228 & 1.17 & 0.17 \\ 
  2000 & Finland & 261 & 0.42 & 0.12 & 267 & 0.89 & 0.15 \\ 
  2000 & France & 253 & 0.04 & 0.04 & 258 & 1.21 & 0.05 \\ 
  2000 & Germany & 250 & -0.64 & 0.04 & 303 & 1.81 & 0.05 \\ 
  2000 & Hungary & 291 & -0.46 & 0.09 & 254 & 1.66 & 0.12 \\ 
  2000 & Italy & 278 & -0.31 & 0.05 & 278 & 0.81 & 0.06 \\ 
  2000 & Netherlands & 129 & 0.46 & 0.08 & 158 & 1.25 & 0.11 \\ 
  2000 & Poland & 209 & -0.31 & 0.04 & 166 & 1.25 & 0.06 \\ 
  2000 & Spain & 318 & -0.27 & 0.05 & 340 & 1.09 & 0.07 \\ 
  2000 & Sweden & 241 & 0.12 & 0.10 & 246 & 1.06 & 0.13 \\ 
  2000 & United Kingdom & 531 & 0.16 & 0.04 & 415 & 1.33 & 0.06 \\ 
  2000 & United States & 222 & -0.43 & 0.02 & 164 & 1.74 & 0.03 \\ 
  2015 & Australia & 1694 & -0.15 & 0.04 & 1235 & 1.28 & 0.05 \\ 
  2015 & Austria & 676 & -0.30 & 0.07 & 705 & 1.53 & 0.10 \\ 
  2015 & Canada & 2215 & 0.20 & 0.03 & 1863 & 0.99 & 0.05 \\ 
  2015 & Denmark & 1013 & 0.09 & 0.08 & 595 & 1.08 & 0.10 \\ 
  2015 & Finland & 575 & 0.07 & 0.08 & 584 & 1.16 & 0.11 \\ 
  2015 & France & 570 & -0.37 & 0.02 & 615 & 1.66 & 0.03 \\ 
  2015 & Germany & 545 & -0.09 & 0.02 & 582 & 1.45 & 0.03 \\ 
  2015 & Hungary & 466 & -0.72 & 0.07 & 589 & 1.79 & 0.09 \\ 
  2015 & Italy & 947 & -0.20 & 0.03 & 1043 & 1.23 & 0.04 \\ 
  2015 & Netherlands & 521 & 0.13 & 0.04 & 525 & 1.24 & 0.06 \\ 
  2015 & Poland & 446 & 0.10 & 0.03 & 448 & 1.24 & 0.05 \\ 
  2015 & Spain & 608 & -0.25 & 0.03 & 703 & 1.24 & 0.04 \\ 
  2015 & Sweden & 527 & -0.20 & 0.06 & 542 & 1.32 & 0.09 \\ 
  2015 & United Kingdom & 1387 & -0.03 & 0.03 & 1388 & 1.15 & 0.04 \\ 
  2015 & United States & 585 & -0.39 & 0.01 & 544 & 1.22 & 0.01 \\ 
   \hline
\end{tabular}
}

The first table shows a description of the sample size and mean score of both top and bottom SES groups for only the first and last time point. One main concern from the planned analysis is that getting the top 90th percentile and bottom 10th percentile would result in a small sample size. The table suggests that we have a reasonable number of respondents to actually estimate gaps accurately. Moreover, we can see that in all instances the bottom SES group has a lower score than all top SES groups.

We see that in some countries like Finland and Sweden the average low SES scores is actually above the average score of 0, suggesting their very equal countries. However, most countries, like Spain, Poland, Hungary and Germany, their average low SES scores are much lower than the country average, which is 0. We also see some countries with major changes from year 2000 to 2015, with, for example, Australia decreasing the low SES gap from 0.22 standard deviations to -0.15, well below the country average. In the next section we take a look at this in a more detailed fashion, considering all years in between.

\subsection{Evolution of the achievement gap}
% Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).

@@

\caption{Evolution of the achievement gap for selected countries}
\end{center}
\end{figure}

We start by look at the achievement for some countries in figure 2. We plot only mathematics for each country and also a quadratic trend spline for \emph{both mathematics and literacy} pooled. Some countries have increased their achievement very strongly. For example, France, Austria and surprisingly Sweden have very steep slopes. France increasde the gap by roughly 0.9 standard deviations, Austria by 0.6 and Sweden by 0.6. This pattern happens similarly for literacy. For example, France has an increase of 0.6. For such a short period of time, the magnitud of these increases are reasonably big.

Given that no one has estimated the evolution of the gap we can't cross-check how other empirical estimations put France at. However, we can take \citet{micklewright} as the closest reference which also finds that France was a low dispersion country in 2000. However, there's no evidence on what happened over time. Luckily, the work of \citet{bernardi2016} did study the relationship of social origin effects (broadly speaking, not in terms of achievement gaps) in France and found that they increased in the last decades.

Other countries have reasonable increases such as Finland and Hungary, with increases of nearly 0.6 and 0.4, respectively. Aside from these countries, there are other cases that have no change at all, specifically, Canada, Netherlands and Spain. Canada excels here not only because the gap has been stable over time, but because it has the smallest gap of all countries presented here. It's nearly 0.5 standard deviations in 2000 and it increased only by 0.2 in 2015. We can't really say Italy has had a significant since it has a sort of wiggly pattern that looks like it cancels itself out.

On the other hand, we do have some countries which show a decrease in the SES achievement gap. United Kingdom decreased by about 0, Poland by -0.4, and Denmark by -0.2. However, the most notable cases are the United States and Germany. These two countries show a very high level of dispersion in the year 2000 with SES gaps of over 2 standard deviations. But in the 15-year time trend both countries reduced the gaps by -0.6 and -0.9 respectively. Their distinctively big gaps in 2000 also show up in the work of \citet{micklewright}. This corroborates the findings of \citet{reardon_portilla}, which found a decreasing gap for kindergarteners. This decline is evident also for 15 year olds, suggesting it might be more of an institutional change rather than a specific grade-level policy.

@@

We find that the initial gap for the U.S in 2000 is 2.17 varying between 2.13 and 2.21. This gap is much higher than found in previous studies, but again, we don't know much previous estimates vary and nobody has really studied the 15-year old achievement gap precisely. All country-level studies pool many studies together and control for age, removing the age-specific gap effect.

Figure 3 below takes a more direct approach and looks at the percentage change from the first and last timepoint available. Each data point has been computed together with it's 50\% and 95\% uncertainty interval \footnote{Each of these uncertainty intervals were computed using a 500-replicate bootstrap}.

@@

\caption{Percentage change in the achievement gap from 2000 to 2015}
\end{center}
\end{figure}

Generally speaking, we see that most countries increased their achievement gap over time. France had an average increase of about 80\% since 2000 varying down to 40\%, whereas Germany had a similar figure but decreasing. Many of the countries we saw before that didn't have a very steep slope, such as Hungary or Australia, in fact had increases of about 40\% of their gap. In contrast, we see that the U.S and Poland had also quite significant decreases of about 40\%. The benefit of presenting these estimates this way is that we can actually asses the uncertainty of each calculation, and we do find that some of these estimates have wide variability. Despite this, most countries show a clear sign of either decreasing or increasing.

Something reassuring is that mathematics and reading (not presented) follow basically the same trend across all countries \footnote{Reading had an even bigger decrease for similar countries, and a much smaller increase. The previos two graphs for reading are available upon request}. This means that the result is not an artifact of chance. However, it's important to disentangle where is this gap originating from. Is this because the top are improving why the bottom decreases? Or is it that the bottom is catching up? Up next we plot the same graph but show the divergent patterns between high/low SES origins.

\subsection{Source of the achievement gap}

@@

\caption{Evolution of the achievement gap for separate SES groups}
\end{center}
\end{figure}

Figure 4 shows the evolution for both Low and High SES groups separetly. The middle line is a summary line showing whether one of the groups is growing/shrinking faster than the other. We find plenty of variation in these patterns. For example, in the United States the top seems to be equalizing much stronger than than the bottom is catching up. The U.K seems to be following the same path as the United States as well. On the other hand, in Poland the gap seems to be closing because the bottom SES group is catching up much faster. The Netherlands shows a similar pattern as the U.S, U.K patterns but upon closer inspection the explanation is very different. We see that the slope is negative (like in the previous two countries) but that's because the low SES group is decreasing at a greater rate than the high SES is decreasing.

Finally, we pay attention to the flat summary lines. These lines should be interpreted with caution because it doesn't mean that the gap is not increasing. Denmark, Belgium, Netherlands, Norway and Spain show a flat line because the gap is growing very little, if at all. On the other hand, Sweden, France, Finland, New Zealand, Austria, among other countries, show a flat line because both groups are distancing themselves at a very similar rate. These results highlight the importance of not only summarizing average achievement gaps. The source of these gaps vary greatly between countries and we see how each of these patterns contibutes to the overall inequality of a country.

Ths source of this gap can be deconstructed further into two broader gaps. We want to know if these gaps have increased because the 90th SES group has distanced itself from the 50th SES group or because the middle has been also distancing itself from 10th SES group. For example, we're interested in searching which gap is bigger between countries, the 90th/50th or the 50th/10th? Moreover, are some of these gaps contributing more to the 90th/10th gap we saw earlier? Up next we plot all the combinations of the 90th/50th/10th gaps and pay attention to the magnitude of each gap in figure 5. Because we visualize three plots for every country, we only plot a selected number of countries that represent the overall patterns from the previous graphs \footnote{United States and Germany show a decrease, France shows a steep increase and Denmark shows no significant change.}.

@@

\caption{Evolution of 90/50/10 SES gaps for selected countries}
\end{center}
\end{figure}

We start off with Denmark. Generally speaking, we find that the 50th/10th gap is contributing more than the 90th/50th gap as the gap is bigger in mathematics. Furthermore, we find, just as in the 90th/10th SES gap, that there isn't a significant decrease \footnote{With the exception of reading in the 50th/10th SES gap which seemed to decrease from nearly 1.5 to 1 standard deviation.}. France shows a similar pattern as Denmark in that the 50th/10th SES gap is the main contributer and that the three gaps have the similar trend.

Germany and U.S, on the other hand, show a very different landscape. For example, Germany, shows that the 90th/10th gap is primarily driven by the 50th/10th SES gap and that the slopes of the 90th/50th gap is much weaker than the other two gaps. This means that top/middle SES groups have shrunk much less than the other groups. The big catching up is actually coming from the middle/bottom SES groups. However, we don't know if this is because the bottom is catching up or that the middle is coming down. Based on the gap plots from above, we can infer that it is because the bottom is catching up rapidly because against the 90th SES group, it is increasing steeply every year.

The U.S shows a different pattern from all other countries. Both the 90th/50th/10th gaps seem to contribute rather similarly to the overall 90th/10th gap (although 50th/10th is marginally higher) but more interestingly, the 90th/50th gap has stagnated in the last 15 years. The top/middle groups have a relatively big gap and it hasn't changed at all. On the other hand, the steep decrease we saw for the U.S is coming exclusively from the 50th/10th gap shrinking. The reason, nonetheless, might be different from the German case, where it's probable that the bottom is catching up. Here, as in the 90th/10th gap, the middle might be 'dumbing' down closer to the low SES group. The big takeaway from the U.S is that the decrease of inequality is coming from the shrinking of the 50/10 gap.

In a more convenient way, we can inspect whether there is a lot of variance in the rate at which top/bottom groups are changing in figure 6.

@@

\caption{Rate at which top/bottom SES groups are catching up for the 90/10 SES gap}
\end{center}
\end{figure}

We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (red panel) but there's also a fair share where low SES is catching up faster, such as the United States and Germany. We see a negative relationship where the faster the low SES catch up the slower the High SES decrease. This is very interesting because it suggests that there's a trend for countries to have the bottom SES groups are climbing upwards, a trait of meritocracy. Equally important, we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel). On the other hand, we do see more ocuntries where the both groups are increasing (top right panel). We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.

But, similar as before, we don't know which group is causing the decrease/increase of the gap in the 90th/50th and 50th/10th gap. For that we plot the evolution of these gaps in figure 7.

@@

Given that we have the data for the average performance for each country and the achievement gap, we can also investigate whether there's an efficiency/equality tradeoff and if it stands over time. One particular topic that has been researched in the past is whether improving the overall cognitive development of a country compromises the distribution of the countries overall performance. For example, in order to improve the average score in, let's say, mathematics, tracking can improve the average score by making children in the higher trackers perform better, raising the overall performance. This, however, can come at a price. If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied with an increase in the overall performance. We plot this relationship in figure 8.

\subsection{Efficiency-equality tradeoff}

@@

{\centering \includegraphics[width=6.5in,height=5in]{figure/avgperformance_inequality-1}

@@

\caption{Relationship between 90/10 achievement gap and average performance in mathematics}
\end{center}
\end{figure}

For the past 15 years we see a very strong correlation suggesting that countries that perform very well in PISA have actually very low achievement gaps (or the other way around) \footnote{We graph all countries-years pooled as well and the relationship is identical}. The correlations are between -0.73 and -0.88. This confirms the theoretical and empirical frameworks suggested by \citet{werfhorst_mijs}. However, what they didn't really go in depth was whether this corelation is strongest for some countries over time. With the data available we can look at the evolution separately for many countries. We plot it up next in figure 9.

@@

\caption{Time trends of achievement gap and average achiement between 2000 and 2015}
\end{center}
\end{figure}

Looking at figure 9, we see a strong correspondence between the two variables. In countries like France we see quite a functional form between both inequality and performance. As average performance plummets, the achievement gap is increasing. This pattern is also seen in Austria, Australia and Germany. The other obvious pattern is the opposite trend. We see countries like Poland, Finland, Canada and less pronounced, Italy, where this pattern is evident. The surprising case here is the United States. We see that as the average performance has been decreasing, the average achievement gap has been decreasing as well. In fact, among the few countries we should find something that resembles this pattern should be the U.S because it doesn't have a strict institutionalized tracking system. However, the explanation behind this pattern is very speculative.

Finally, the last notable pattern is that in some countries the average performance is so high that it actually overperforms the achievement gap. Only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality, which coincidentially, are countries that have the lower leven of inequalities among the ones presented here.
